DNA Methylation Hierarchical Variance Model\nStan Results

### 2021-07-29 Paper_Figures.Rmd
updated analysis to exclude SNPs, new gene sets for neuron marker and cardiac progenitor genes

# remove the corresponding rows from the results dataframes
FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
mu_full <- mu_full[-union(QCinds,XYinds),]
nu_full <- nu_full[-union(QCinds,XYinds),]
sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
lp_full <- lp_full[-union(QCinds,XYinds),]
CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
```

@@

@@

# plot violins of: all genes, COSMIC, AGO
perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n)) %>% mutate(score_median = 2*score_median)
genes_dbs <- rbind(perm_temp %>% mutate(group="all genes"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (essentialgenes %>% pull(X1))) %>% mutate(group="DEPMAP")
              )
# Fig 4C: plot distribution
pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP"))) %>%
  #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
  ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
  scale_x_continuous(limits=c(-5,7), breaks = c(-4, -2,0, 2, 4, 6)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
# # add hashmarks for each gene (doesn't look great)
# marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
pDB
# ggsave("database_violins.svg")
ggsave("database_boxplot_depmap.svg")

@@

# stats
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
```

\newpage
\subsection{Oncogenes vs Tumor suppressors}
Based on list of top 20 genes mutated in CRC, manually labelled as oncogene or tumor-suppressor:

| Oncogene | Tumor Suppressor |
|----------|------------------|
| KRAS     | TP53             |
| MYC      | APC              |
| BRAF     | KLF4             |
| EGFR     | FBXW7            |
| CDK8     | LRP1B            |
| PIK3CA   | PTEN             |
| NRAS     | FAT4             |
| GNAS     | ARID1A           |
| SRC      | AMER1            |
| CTNNB1   | SMAD4            |

```{r Onco_vs_TSG, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA, fig.fullwidth=TRUE, dev = "png"}

plotGeneScore <- function(geneStr,norm) {
  cat("\n")
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
}

cat("\nOncogenes:")
plotGeneScore("KRAS",0.25)
plotGeneScore("MYC",0.21)
plotGeneScore("BRAF",0.5)
plotGeneScore("EGFR",0.5)
plotGeneScore("CDK8",0.5)
plotGeneScore("PIK3CA",0.5)
plotGeneScore("NRAS",0.5)
plotGeneScore("GNAS",0.5)
plotGeneScore("SRC",0.5)
plotGeneScore("CTNNB1",0.5)

cat("\nTumor Suppressors:")
plotGeneScore("TP53",0.25)
plotGeneScore("APC",0.21)
plotGeneScore("KLF4",0.5)
plotGeneScore("FBXW7",0.5)
plotGeneScore("LRP1B",0.5)
plotGeneScore("PTEN",0.5)
plotGeneScore("FAT4",0.5)
plotGeneScore("ARID1A",0.5)
plotGeneScore("AMER1",0.5)
plotGeneScore("SMAD4",0.5)

>>>>

# use MethylToSNP package to detect likely SNPs
# from paper: LaBarre et al. 2019, Epigenetics & Chromatin)
# methyltosnp_results file should exist from running StanAnalysis.Rmd
snps <- read.csv(here("methyltosnp_results.csv"))
rownames(snps) <- snps[,1]
snps<-snps[,-1]
SNPinds <- which(FullAnnotation$IlmnID %in% rownames(snps))

remove_inds = union(QCinds, union(XYinds, SNPinds))

# remove the corresponding rows from the results dataframes
FullAnnotation <- FullAnnotation[-remove_inds,]
mu_full <- mu_full[-remove_inds,]
nu_full <- nu_full[-remove_inds,]
sigmaP_full <- sigmaP_full[-remove_inds,]
sigmaPT_full <- sigmaPT_full[-remove_inds,]
sigmaT_full <- sigmaT_full[-remove_inds,]
sigmaE_full <- sigmaE_full[-remove_inds,]
lp_full <- lp_full[-remove_inds,]
CpGscore_full <- CpGscore_full[-remove_inds,]
```

@@

# additional non-relevant gene sets from MSigDB (LEIN_NEURON, CARDIAC_PROGENITOR)
# MSigDB: C2: LEIN_NEURON_MARKERS
neuron <- read_delim("/Users/kevinmurgas/Documents/Data+ project/geneset_neuron.gmt", delim="\t") %>% colnames()
# MSigDB: C2: WP_CARDIAC_PROGENITOR_DIFFERENTIATION
cardprog <- read_delim("/Users/kevinmurgas/Documents/Data+ project/geneset_cardprog.gmt", delim="\t") %>% colnames()

@@

# plot violins of: all genes, COSMIC, AGO
perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n)) %>% mutate(score_median = 2*score_median)
genes_dbs <- rbind(perm_temp %>% mutate(group="all genes"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (essentialgenes %>% pull(X1))) %>% mutate(group="DEPMAP"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% neuron) %>% mutate(group="neuron"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% cardprog) %>% mutate(group="cardprog")
              )
# Fig 4C: plot distribution
pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP", "neuron", "cardprog"))) %>%
  #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
  ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
  scale_x_continuous(limits=c(-5,7), breaks = c(-4, -2,0, 2, 4, 6)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
# # add hashmarks for each gene (doesn't look great)
# marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
# pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
pDB
# ggsave("database_violins.svg")
ggsave("database_boxplot_depmap.svg")

@@

# stats
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="neuron"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="cardprog"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
```

### 2021-07-29 StanAnalysis.Rmd
updated analysis to exclude SNPs, new gene sets for neuron marker and cardiac progenitor genes

\subsection{Filter CpG sites}
A filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit), and CpG sites mapping to sex chromosomes X and Y are removed as well to eliminate sex-specific bias.

@@

# remove the corresponding rows from the results dataframes
FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
mu_full <- mu_full[-union(QCinds,XYinds),]
nu_full <- nu_full[-union(QCinds,XYinds),]
sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
lp_full <- lp_full[-union(QCinds,XYinds),]
CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
```

>>>>

\subsection{Filter CpG sites}
Three filters are applied to remove CpG sites from subsequent analysis:
1. Sites with poor model convergence: Rhat > 1.1 in lp (log-posterior likelihood of entire site fit),
2. Sites mapping to sex chromosomes X and Y,
3. Sites with likely SNPs (determined via MethylToSNP algorithm from LaBarre 2019, *Epigenetics & Chromatin*)

8740, 19632, and 377 sites were identified in each respective filter, for total union of 28557 CpG sites to remove (out of 866091 total CpG sites)

@@

# detect likely SNPs with MethylToSNP package
# from paper: LaBarre et al. 2019, Epigenetics & Chromatin)
if ( file.exists(here("methyltosnp_results.csv")) ) {
  # if file already exists then load it
  snps <- read.csv(here("methyltosnp_results.csv"))
  rownames(snps) <- snps[,1]
  snps<-snps[,-1]
} else {
  # otherwise, run the MethylToSNP algorithm and save
  library(MethylToSNP)
  library(minfiDataEPIC)
  data(SNPs.147CommonSingle)
  data_beta = FullAnnotation[,10:57]
  rownames(data_beta) <- FullAnnotation$IlmnID
  snps <- MethylToSNP(data_beta, SNP=SNPs.147CommonSingle) %>% filter(confidence>0.5) # use suggested cutoff of 0.5
  write.csv(snps, file = here("methyltosnp_results.csv"))
}
SNPinds <- which(FullAnnotation$IlmnID %in% rownames(snps))

remove_inds = union(QCinds, union(XYinds, SNPinds))

# remove the corresponding rows from the results dataframes
FullAnnotation <- FullAnnotation[-remove_inds,]
mu_full <- mu_full[-remove_inds,]
nu_full <- nu_full[-remove_inds,]
sigmaP_full <- sigmaP_full[-remove_inds,]
sigmaPT_full <- sigmaPT_full[-remove_inds,]
sigmaT_full <- sigmaT_full[-remove_inds,]
sigmaE_full <- sigmaE_full[-remove_inds,]
lp_full <- lp_full[-remove_inds,]
CpGscore_full <- CpGscore_full[-remove_inds,]
```
